Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-09 paper.tex
Minor updates

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools such as
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
features, links with domain-specific tool repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide-array of
workflows have been implemented using these pipelines and are now
shared across neuroimaging groups world-wide. These domain-specific
engines represent a tremendous opportunity for science gateways to
leverage existing tools and applications. They nicely complement
e-Science engines that are more oriented towards the exploitation of
distributed computing infrastructures, in particular grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipelines.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including complete
frameworks such as Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these frameworks
and instances include one or several workflow engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and instead focus on the execution of simple tasks.

@@

Specific features may also be available in science gateways, such as
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

In this paper, infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may assume specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler,
the presence of a file catalog, etc. In the analysis presented below,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

  workflow to be executed. The interaction has two aspects:

@@

  SHIWA initiative. In SHIWA, workflow conversion is done through an
  intermediate representation, the Interoperable Workflow Intermediate
  Representation~\cite{plankensteiner-montagnat-etal:2011}, which
  allows to convert among $n$ workflow languages using $2n$
  interactions instead of $n^2$.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}. In the SHIWA
pool, agents can wrap different types of workflow engines to execute
workflows expressed with different languages.

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language that provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}. Tavaxy~\cite{Abouelhoda2012}
allows to import, merge and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration complexity, robustness,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. Adding a new type
  of workflow engine allows to execute more workflows in the science
  gateway.
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

>>>>

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools such as
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and others. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

At the same time, toolboxes have been emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), RPPL~\cite{1174106} and
PMP (Poor Man's Pipeline~\cite{Ad-DabbaghY2006}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular
clusters\closednote{Tristan}{Reverted from Marc's 'cluster
  scheduler'}{}{}, to execute tasks at a high throughput. Some of
these tools also support advanced features such as provenance
tracking, \closednote{Naj}{should we not mention PMP which is used for
  CIVET?}{Tristan}{Done, as discussed.}  redundancy detection across
analyses to avoid re-computation, and so on. A wide array of workflows
have been implemented using these pipelines and are now shared across
neuroimaging groups world-wide, which represents a tremendous
opportunity for science gateways to leverage. Domain-specific engines
nicely complement e-Science systems that are more oriented towards the
exploitation of distributed computing infrastructures, in particular
grids and clouds.

In this paper, a
\emph{workflow engine} (also abbreviated \emph{engine}) is a piece of
software that submits interdependent computing \emph{tasks} to an
\emph{infrastructure} (local server, cluster, grid or cloud) based on
a workflow description (a.k.a \emph{workflow}) and using input data
that may consist of files, database entries or simple parameter
values. Although simplistic, this definition covers both e-Science workflow systems and
domain-specific pipeline systems.\closednote{Marc-e}{To remain consistent with above descriptions}{Tristan}{yes!} Some workflow engines, usually e-Science
ones, may transfer data across the infrastructure, and others, usually
domain-specific ones, may leave this role to external processes. Workflows
may be expressed in any language, including high-level XML or JSON
dialects such as Scufl or Hyperflow, and low-level scripting languages
such as Bash.

@@

Various science gateways have been developed, including frameworks
such as Apache Airavata~\cite{marru2011apache}, the Catania Science
Gateway Framework~\cite{ardizzone2012decide} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide} or
as standalone systems~\cite{SHER-14,GLAT-13}. Most of these systems
include one or several workflow engines.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. 
%Some gateways may not even use any
%workflow engine and instead focus on the execution of simple tasks.

@@

Specific features may also be available, for instance
data vizualisation and quality control, workflow edition, debugging
instruments, social tools among users, etc.

@@

Infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may require specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler,
the presence of a file catalog, etc. In the analysis presented below,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

  workflow to be executed. The interaction has two steps:

@@

  SHIWA initiative.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored\closednote{sorina}{VIP and Moteur also share the
  WorkflowsDB}{Tristan}{Yes, you are right. This is an example where
  such abstract architectures may not completely apply in real
  life. See comment about ``blended'' architectures in the
  discussion. In case of VIP, the sharing of the DB is more an
  implementation detail rather than a strong architectural pattern: we
  could avoid sharing the DB if we wanted.}.  In this model, task and
data control are both initiated from the science gateway. Interactions
\texttt{b} and \texttt{c$_2$} are initiated from the workflow engine
while \texttt{c$_1$} comes from other parts of the science gateway,
for instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}), in the Distributed
application runtime environment
(DARE~\cite{maddineni2012distributed}), in
Galaxy~\cite{goecks2010galaxy}, and in the LONI Pipeline
Environment~\cite{dinov2009efficient}. Note that tightly integrated
architectures may provide advanced workflow edition features which are
not covered by our analysis.
%The Catania Science Gateway Framework would be a good illustration here (workflow engine is a Java portlet), but I can't find a suitable figure about it.
%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

The pool model was implemented in the SHIWA pool~\cite{ROGE-13}
diagrammed in Figure~\ref{fig:shiwa-pool-architecture}, where agents
can wrap different types of workflow engines to execute workflows
expressed with different languages.

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language that provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows to convert
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
allows to import, merge and run Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.
\closednote{Tristan}{Add description of a real system}{Tristan}{Would
  require to reproduce a figure from another paper. The concept seems
  simple enough to not require a figure.}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation} using 5 main criteria:
integration complexity, robustness, extensibility, scalability and
other specific features. Criteria break down to specific metrics where
\emph{lower value indicates better performance}. For each criterion, a
global score is computed by summing up the individual metrics. We
ensure that the different metrics in a criterion measure similar
entities so that they can be summed. The criteria and metrics are
explained hereafter.

@@

\emph{Extensibility} measures the difficulty to replace or add
elements in the architecture. It is determined as the number of interactions and
components that need to be modified when a new element is
added. Modification of a component is required when its code needs to
be updated or recompiled (science gateway or workflow service), or
when a new piece of software has to be installed (infrastructure
only). Modification of an interaction is deemed necessary when the
parameters involved in this interaction are modified.  Extensibility
breaks down in 4 metrics depending on the type of element that has to
be added or replaced:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item New engine type (\texttt{E$_1$}): number of interactions or
  components to modify to integrate a new type of workflow engine in
  the architecture. Workflow engines belong to different types when
  they cannot be invoked using the same interface. 
\item New engine version (\texttt{E$_2$}): number of interactions or
  components to modify to integrate a new version of a workflow engine
  in the architecture, assuming that another version of the same
  engine type is already available. Different versions of a workflow
  engine share the same interface, i.e. they can be invoked using the
  same software. When this is not the case, the different versions are
  considered as different engine types. Components are not treated
  equally regarding engine version updates. Components whose only
  function is to host the workflow engine, i.e. workflow services and
  agents, are not counted in E$_2$ because updates in such components
  are assumed to be straightforward. On the contrary, modifications of
  components that have other functions than wrapping the engine,
  i.e. science gateway and infrastructure, are counted in E$_2$
  because updates in these components require more effort, e.g. new
  release of the science gateway, gaining administrative privileges on
  the infrastructure, etc.
\item New workflow (\texttt{E$_3$}): adding a new workflow is a very
  common operation that does not require modifying software components
  or interactions. We measure the difficulty to integrate a new
  workflow by counting the number of interactions required
  to integrate a new workflow in the architecture, assuming
  that the engine type and version required to execute this workflow
  are already available.
\item New infrastructure (\texttt{E$_4$}): number of interactions or
  components to modify to integrate a new type of infrastructure in
  the architecture. Adding a new infrastructure allows to provide more
  computing or storage power, to access specific types of resources
  (e.g., GPUs, clouds), or to enforce execution policies (e.g.,
  constrain data to remain in a particular network domain).
%\item New science gateway: \todo{Not sure if this makes sense}.
\end{itemize}
The 4 metrics are summed to obtain a global index that measures the
difficulty to extend the architecture. \closednote{Naj}{I would rather compare each metric separately, this woudl also give a sense of how different architectures compare across metrics}{Tristan}{Metrics are compared separately too, see Table 2 and discussion.} Note that some extensions may
involve several metrics in practice. For instance, adding a new type
of engine may help integrate new infrastructures when interactions
\texttt{b} and \texttt{c$_2$} are already present for the new engine
and infrastructure.  \closednote{Rafael}{Adding a new engine type may
  also allow run in different infrastructures. Maybe it is worth to
  put a sentence to clarify that the metric account for it
  separately.}{Tristan}{It will not *allow* because interaction c1
  would still need to be developed. But it will definitely facilitate
  it. I added a note on this after the description of the metrics.}

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the missing scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
realistically be implemented. \closednote{Naj}{how do you metricize
  "nonesensical amount fo effors?"}{Tristan}{You don't :) Rephrased to
  'cannot realistically be implemented'. At some point it's a matter
  of common sense and experience to draw the line between what is
  easily implemented and what is more tricky. I would like to avoid
  'impossible' because nothing is impossible to implement if enough
  effort is put.} Four different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.  \closednote{Rafael}{$S_2$
    would be 2 if the paradigm does not allow to distribute the
    orchestration? e.g., workflow executions implemented within MPI
    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
    it is very difficult to define ``does not allow to''. For
    instance, why couldn't you hack Pegasus PMC so much that it
    becomes possible to distribute workflow executions in several MPI
    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
    prefer to leave it like this, unless you have a suggestion to
    reflect this in the text.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complicate the task
  scheduling problem by introducing additional software layers or
  creating tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.
